%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% University of Western Ontario Thesis Template
% By: Justin Quinn Veenstra, 2010
% With thanks to Mr. (soon to be Dr.) Will Robertson.


\documentclass[12pt,twoside]{report}
%% Decomment next line to use PostScript fonts
%%\UsePackage{times}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                                                    %%
%%                    ***   I M P O R T A N T   ***                   %%
%%                                                                    %%
%% Fill in the following fields with the required information:        %%
%%  - \department{...}  name of the graduate department               %%
%%  - \degree{...}      name of the degree obtained                   %%
%%  - \author{...}      name of the author                            %%
%%  - \title{...}       title of the thesis                           %%
%%  - \gyear{...}       year of graduation                            %%
%%  - \super{...}    supervisor
%%  - \firstname, \middlename, \lastname... there is additional documentation by the actual fields, so I'll leave it at that
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{appendix}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage[byname]{smartref}
%\usepackage{hyperref} %comment out for hardcopy
\usepackage{txfonts}
\usepackage{tocloft}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{color}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{graphicx}
\usepackage{glossaries}
\usepackage{bm}
\usepackage{mathtools}
\usepackage{enumerate}
\usepackage{tikz}
\usetikzlibrary{arrows}
\usetikzlibrary{bayesnet}
\usepackage{caption}
\usepackage{url}
\usepackage{rotating}
\usepackage[noabbrev]{cleveref}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{xcolor}
\usepackage{setspace} 

\usepackage[colorinlistoftodos]{todonotes}
\usepackage{changes}

\newcommand{\Normal}{\operatorname{Normal}}
\newcommand{\Gmma}{\operatorname{Gamma}}
\newcommand{\Lognormal}{\operatorname{Lognormal}}
\newcommand{\needscite}{\colorbox{red}{Needs Citation}}

\makeatletter
\numberwithin{figure}{chapter}
\newenvironment{acknowledgements}%
{\clearemptydoublepage
	\begin{center}
		\section*{Acknowledgements}
	\end{center}
	\begingroup
}{\newpage\endgroup}

\newenvironment{dedication}%
{\clearemptydoublepage 
	\begin{center}
		\section*{Dedication}
	\end{center}
	\begingroup
}{\newpage\endgroup}

\newenvironment{preliminary}%
{\pagestyle{plain}\pagenumbering{roman}}%
{\pagenumbering{arabic}}

\addtoreflist{chapter}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}

\newenvironment{proof}[1][Proof]{\begin{trivlist}
		\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
\newenvironment{definition}[1][Definition]{\begin{trivlist}
		\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
\newenvironment{example}[1][Example]{\begin{trivlist}
		\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
\newenvironment{remark}[1][Remark]{\begin{trivlist}
		\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}

\newcommand{\qed}{\nobreak \ifvmode \relax \else
	\ifdim\lastskip<1.5em \hskip-\lastskip
	\hskip1.5em plus0em minus0.5em \fi \nobreak
	\vrule height0.75em width0.5em depth0.25em\fi}

% Default values for title page.

%% To produce output with the desired line spacing, the argument of
%% \spacing should be multiplied by 5/6 = 0.8333, so that 1 1/2 spaced
%% corresponds to \spacing{1.5} and double spaced is \spacing{1.66}.
\def\normalspacing{1.25} % default line spacing


%% Define the "thesis" page style.
\if@twoside % If two-sided printing.
\def\ps@thesis{\let\@mkboth\markboth
	\def\@oddfoot{}
	\let\@evenfoot\@oddfoot
	\def\@oddhead{
		{\sc\rightmark} \hfil \rm\thepage
	}
	\def\@evenhead{
		\rm\thepage \hfil {\sc\leftmark}
	}
	\def\chaptermark##1{\markboth{\ifnum \c@secnumdepth >\m@ne
			Chapter\ \thechapter. \ \fi ##1}{}}
	\def\sectionmark##1{\markright{\ifnum \c@secnumdepth >\z@
			\thesection. \ \fi ##1}}}
\else % If one-sided printing.
\def\ps@thesis{\let\@mkboth\markboth
	\def\@oddfoot{}
	\def\@oddhead{
		{\sc\rightmark} \hfil \rm\thepage
	}
	\def\chaptermark##1{\markright{\ifnum \c@secnumdepth >\m@ne
			Chapter\ \thechapter. \ \fi ##1}}}
\fi

\pagestyle{thesis}
% Set up page layout.
\setlength{\textheight}{9in} % Height of the main body of the text
\setlength{\topmargin}{-.5in} % .5" margin on top of page
\setlength{\headsep}{.5in}  % space between header and top of body
\addtolength{\headsep}{-\headheight} % See The LaTeX Companion, p 85
\setlength{\footskip}{.5in}  % space between footer and bottom of body
\setlength{\textwidth}{6.25in} % width of the body of the text
\setlength{\oddsidemargin}{.25in} % 1.25" margin on the left for odd pages
\setlength{\evensidemargin}{0in} % 1.25"  margin on the right for even pages

% Marginal notes
\setlength{\marginparwidth}{.75in} % width of marginal notes
\setlength{\marginparsep}{.125in} % space between marginal notes and text

% Make each page fill up the entire page. comment this out if you
% prefer. 
\flushbottom

\setcounter{tocdepth}{1} % Number the subsubsections 
\def\normalspacing{1.25} % default line spacing

\newcommand\isco[1]{%
	\edef\@tempa{#1}%
	\def\@tempb{}%
	\ifx\@tempa\@tempb
	\else \\\underline{Co-Supervisor:}\vspace{0.35in}\\\dots\dots\dots\dots\dots\dots\dots\\{#1}\\
	\fi
}

\newcommand\isjoint[1]{%
	\edef\@tempa{#1}%
	\def\@tempb{}%
	\ifx\@tempa\@tempb
	\else \\\underline{Joint Supervisor:}\vspace{0.35in}\\\dots\dots\dots\dots\dots\dots\dots\\{#1}\\
	\fi
}

\newcommand\isalt[1]{%
	\edef\@tempa{#1}%
	\def\@tempb{}%
	\ifx\@tempa\@tempb
	\else \\\underline{Alternate Supervisor:}\vspace{0.35in}\\\dots\dots\dots\dots\dots\dots\dots\\{#1}\\
	\fi
}

\newcommand\isdefinedsig[1]{%
	\edef\@tempa{#1}%
	\def\@tempb{}%
	\ifx\@tempa\@tempb
	\else \\ \dots\dots\dots\dots\dots\dots\dots\\{#1}\\
	\fi
}
\newcommand\isdefinedspinetitle[1]{%
	\edef\@tempa{#1}%
	\def\@tempb{}%
	\ifx\@tempa\@tempb
	\else (Spine title: #1)\\
	\fi
}
\newcommand\coauthor[1]{%
	\edef\@tempa{#1}%
	\def\@tempb{}%
	\ifx\@tempa\@tempb
	\else \newpage \Large Co-Authorship Statement\normalsize\\\indent\\#1\\
	\fi
}

\newcommand\acknowlege[1]{%
	\edef\@tempa{#1}%
	\def\@tempb{}%
	\ifx\@tempa\@tempb
	\else \newpage \Large Acknowledgements\normalsize\\\indent\\#1\newpage
	\fi
}

%\renewcommand{\appendixtocname}{\Huge \textbf{List of Appendices} \normalsize}
\newcommand{\blank}{\hspace{-2mm}}
\newcommand{\super}{Dr. Daniel J. Lizotte} %supervisor
\newcommand{\superj}{} %joint supervisor, if there is one, leave blank if not (lbin)... only one of the three.
\newcommand{\superc}{} %co-supervisor, if there is one, leave blank if not (lbin)
\newcommand{\supera}{} %alternate supervisor, if there is one, leave blank if not (lbin)
\newcommand{\sco}{Committee Member 1}  %member of supervisory committee
\newcommand{\sct}{Committee Member 2}  %other member of supervisory committee (lbin)
\newcommand{\sch}{Committee Member 3}  %other member of supervisory committee (lbin)
\newcommand{\examo}{Examiner 1}  %examining committee (up to four, if less leave blank)
\newcommand{\examt}{Examiner 2}
\newcommand{\examth}{Examiner 3}
\newcommand{\examf}{}
\newcommand{\department}{Epidemiology and Biostatistics}
\newcommand{\degree}{Doctor of Philosophy}
\newcommand{\firstname}{Athanasios}
\newcommand{\middlename}{Demetrios}
\newcommand{\lastname}{Pananos}
%\renewcommand{\author}[1]{\ifx\empty#1\else\gdef\@author{#1}\fi} 
\newcommand{\authorname}{{\firstname} {\middlename} {\lastname}}
\newcommand{\titl}{Bayesian Pharmacokinetic Models for Inference and Optimal Sequential Decision Making with Applications in Personalized Medicine}
\newcommand{\spinetitle}{Bayesian Pharmacokientic Models with Applications in Personalized Medicine}%only if the above is more than 60 characters
\newcommand{\thesisformat}{Integrated Article} %or Integrated Article
\newcommand{\gyear}{\number\year}
\newcommand{\makecoauthor}{
	%Type information about coauthorship here/
	I would like to acknowlege my imaginary friend, Jummi for doing all the work. 
}
\newcommand{\makeacknowlege} {
	
	First, thank you to Dr.\ Dan Lizotte.  Dan has been an incredible mentor. So far as I can tell, Dan percieves me not as a trainee but as a colleague.  He has been patient, supportive, and kind.  I envy all future students who have the privellage to work with him.
	\newline
	\newline
	Thank you to my committe  for taking the time to read my thesis and provide edits to various components over the last 5 years.
	\newline
	\newline
	Thank you to Gary, who has been the best friend I have had. Our near daily statistical chats over coffee are something I look back on fondly.  For all the times we thought ``we should just quit'' -- and there were many -- I'm glad we didn't.
	\newline
	\newline
	Thank you to Dr.\ Simon Bonner, who I credit with cultivating the applied statistician I am today.  Much of my own statistical philosophy came from your Generalized Linear Models course.  It was a foundational course, in more ways than one, and represents a sort of line in the sand from which I started to call myself a statsitician propper.
}
\newcommand{\listappendixname}{List of Appendices}
\newlistof{myappendices}{app}{\listappendixname}
\newcommand{\myappendices}[1]{%
	\addcontentsline{app}{myappendices}{#1}\par}

\renewcommand{\maketitle}
{\begin{titlepage}
		\setcounter{page}{1}
		%% Set the line spacing to 1 for the title page.
		%\begin{spacing}{1} 
		\begin{large}
			\begin{center}
				\mbox{}
				\vfill
				{\MakeUppercase{\titl}}\\
				\isdefinedspinetitle{\spinetitle}
				(Thesis format: \thesisformat)\\
				\vfill
				by \\
				\vfill
				{\firstname} {\middlename} \underline{\lastname}\\
				\vfill
				Graduate Program in {\department}\\
				\vfill
				A thesis submitted in partial fulfillment\\
				of the requirements for the degree of\\
				\degree\\
				\vfill
				The School of Graduate and Postdoctoral Studies\\
				The University of Western Ontario\\
				London, Ontario, Canada\\
				\vfill
				{\copyright} {\authorname} {\gyear}  \\
				\vspace*{.2in}
			\end{center}
		\end{large}
		%   \end{spacing}
\end{titlepage}

}%\maketitle

\newcommand{\makecert}{
\setcounter{page}{2}
\vfill
\begin{center}
	\large
	THE UNIVERSITY OF WESTERN ONTARIO\\
	School of Graduate and Postdoctoral Studies\\
	\vfill
	\textbf{CERTIFICATE OF EXAMINATION}
\end{center}

\vfill
\begin{table}[ht]
	\begin{minipage}[t]{0.5\linewidth} %tabular instead?
		\begin{tabular}{l}
			\underline{Supervisor:}\vspace{0.35in}
			\isdefinedsig{\super}
			\isco{\superc}
			\isjoint{\superj}
			\isalt{\supera}
			\\
			\underline{Supervisory Committee:}\vspace{0.35in}
			\isdefinedsig{\sco}\vspace{0.15in}
			\isdefinedsig{\sct}\vspace{0.15in}
			\isdefinedsig{\sch}
		\end{tabular}
		\vfill
	\end{minipage}
	\hspace{0.5in}
	\begin{minipage}[t]{0.5\linewidth}
		\begin{tabular}{l}
			\underline{Examiners:} \\\vspace{.5cm}
			\isdefinedsig{\examo}\\
			\isdefinedsig{\examt}\\
			\isdefinedsig{\examth}\\
			\isdefinedsig{\examf}
		\end{tabular}
		\vfill
	\end{minipage}
	\vfill
\end{table}
\vfill
\begin{center}
	The thesis by \\ \vfill
	\textbf{\firstname{} \middlename{} \underline{\lastname}}\\
	\vfill
	entitled:\\\vfill
	\textbf{\titl}\\\vfill
	is accepted in partial fulfillment of the \\
	requirements for the degree of\\
	\degree\\
\end{center}
\begin{table}[ht]
	\begin{minipage}[t]{0.5\linewidth}
		\begin{tabular}{l}
			\dots\dots\dots\dots\dots\\
			Date
		\end{tabular}
	\end{minipage}
	\hspace{0.5in}
	\begin{minipage}[t]{0.5\linewidth}
		\begin{tabular}{l}
			\dots\dots\dots\dots\dots\dots\dots\dots\dots\dots\\
			Chair of the Thesis Examination Board
		\end{tabular}
	\end{minipage}
\end{table}

}

% Defined commands
\newcommand{\rhat}{\widehat{R}}


\makeatother
\begin{document}

%% ***   NOTE   ***
%% You should put all of your '\newcommand', '\newenvironment', and
%% '\newtheorem's (in other words, all the global definitions that you
%% will need throughout your thesis) in a separate file and use
%% "\input{filename}" to input it here.


%% This sets the page style and numbering for preliminary sections.
\begin{preliminary}
	
	%% This generates the title page from the information given above.
	\maketitle
	\addcontentsline{toc}{chapter}{Certificate of Examination}
	\makecert
	\newpage
	%\addcontentsline{toc}{chapter}{Co-Authorship Statement}
	%\coauthor{\makecoauthor}  %comment this out if none
	%\newpage
	\addcontentsline{toc}{chapter}{Acknowlegements}
	\acknowlege{\makeacknowlege}	%as above
	\addcontentsline{toc}{chapter}{Abstract}
	\Large\begin{center}\textbf{Abstract}\end{center}\normalsize
	%%  ***  Put your Abstract here.   ***
	%% (150 words for M.Sc. and 350 words for Ph.D.)
	
	Patients can react to a drug differently just by virtue of being different people, and this between-patient variability in drug response is an obstacle to optimal treatment.  Pharmacokinetic modelling offers one approach to studying drug response, often with covariate focused dose adjustment criteria being reported along side pharmacokinetic modelling.  However, excess variation in concentrations continues to be reported despite the use of these criteria, bringing into question optimal dosing strategies for some drugs.  This thesis provides methods for creating Bayesian pharmacokinetic models for two purposes: inference into the effects of covariates on concentrations and optimal sequential decision making for dose size.   The thesis addresses three objectives: To compare existing approaches to fitting Bayesian models with recent advancements in pursuit of fitting population pharmacokinetic models, to develop a framework for evaluating the benefits of collecting additional information for use in personalization, and to demonstrate how academic personalized medicine researchers can use all data available to them to study effects of clinical variables on pharmacokinetics. To this end, the thesis makes three research contributions. First, a simulation study demonstrating that inferences using popular inference methods in pharmacokinetic research can lead to different and poorer calibrated decisions as compared to newer inference methods.  The model presented in the simulation study was developed using a specific parameterization achieved through non-dimensionalization of the differential equation governing the mass transit of the drug and enables more reliable inference by sampling using Hamiltonian Monte Carlo as compared to a standard parameterization. Second, a unified framework for the development and simulation based evaluation of personalization based on pharmacokinetic modelling combined with dynamic treatment regimes. Lastly, a demonstration of how investigators can fit Bayesian pharmacokinetic models with the aim of accurate modelling of pharmacokinetics and exploration of novel variables using data from heterogeneous sources.  These contributions provide methodologies do address two central goals of personalized medicine -- identification of factors driving between patient variability in drug response, and selection of an optimal dose --  and can enable a richer set of personalized decisions to be made.
	
	\vfill
	\textbf{Keywords:} Bayesian Statistics, Pharmacokinetics, Hierarchical Model, Personalized Medicine
	\newpage
	\tableofcontents\newpage
	\newpage
	\addcontentsline{toc}{chapter}{List of Figures}
	\listoffigures
	\newpage
	\addcontentsline{toc}{chapter}{List of Tables}
	\listoftables\newpage
%	\addcontentsline{toc}{chapter}{List of Appendices}
%	\listofmyappendices\newpage
	%\addcontentsline{toc}{chapter}{List of Abbreviations, Symbols, and Nomenclature}
	%\large List of Abbreviations, Symbols, and Nomenclature \normalsize
	%\newpage
\end{preliminary}
%% End of the preliminary sections: reset page style and numbering.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                                                    %%
%%                    ***   I M P O R T A N T   ***                   %%
%%                                                                    %%
%% Put your Chapters here; the easiest way to do this is to keep each %%
%% chapter in a separate file and \include all the files right here.  %%
%% Note that each chapter file should start with the line             %%
%% "\chapter{ChapterName}".  Note that using "\include" instead of    %%
%% "\input" makes each chapter start on a new page.                   %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% As an example...

\doublespacing
\input{chapters/chapter_1/introduction}		
\input{chapters/chapter_2/sections/differential_equations}
\input{chapters/chapter_2/sections/bayesian_statistics}
\input{chapters/chapter_2/sections/pharmacokinetics}
\input{chapters/chapter_2/sections/reinforcement_learning}


\include{chapters/chapter_3/lit_review}

\input{chapters/chapter_4/sections/introduction}
\input{chapters/chapter_4/sections/background}
\input{chapters/chapter_4/sections/methods}
\input{chapters/chapter_4/sections/results}
\input{chapters/chapter_4/sections/conclusion}
\input{chapters/chapter_4/sections/appendix}


\input{chapters/chapter_5/sections/introduction}
\input{chapters/chapter_5/sections/background}
\input{chapters/chapter_5/sections/framework}
\input{chapters/chapter_5/sections/results}
\input{chapters/chapter_5/sections/discussion_and_conclusion}
\input{chapters/chapter_5/sections/appendix}



\input{chapters/chapter_6/sections/introduction}
\input{chapters/chapter_6/sections/background}
\input{chapters/chapter_6/sections/methods}
\input{chapters/chapter_6/sections/results}
\input{chapters/chapter_6/sections/discussion}

\include{chapters/chapter_7/discussion}


%% This adds a line for the Bibliography in the Table of Contents.
\addcontentsline{toc}{chapter}{Bibliography}
%% ***   Set the bibliography style.   ***
\bibliographystyle{plain} % (change according to your preference)
%%% ***   Set the bibliography file.   ***
\bibliography{bibliographies/thesis_bib}
%% ***   NOTE   ***
%% If you don't use bibliography files, comment out the previous line
%% and use \begin{thebibliography}...\end{thebibliography}.  (In that
%% case, you should probably put the bibliography in a separate file
%% and \include or \input it here).

%Appendices.
\begin{appendices}
	% As ab example...
	%\include{appendixa}
\end{appendices}


\end{document}